\documentclass[]{book}

\input{header.tex}

%opening
\title{High dimensional portfolios - estimation and uncertainty}
\author{Erik Thorsén}

\begin{document}

\maketitle

\section*{Acknowledgements}

\newpage
\section*{List of papers}

\tableofcontents
%%%%%% ------------------------------------------------------------------------
\chapter[Introduction]{Introduction - making decisions and allocations}\label{ch:intro}
%%%%%% ------------------------------------------------------------------------
% General introduction to why portfolio theory and allocations are hard
% ca 2pages
The methods and papers derived in this thesis relies on the fact that investors want to diversify a lot. 
There should be many assets in your portfolio so that whenever one asset might go poorly or even bust, its not going to affect your wealth. 
Its a small portion of your portfolio, e.g. your collection of assets. 
Whenever you are sensitive to losses, or risk averse you should invest in many assets so that whenever .


A lot of diversification might not be for everyone. 
In one of the interviews with Warren Buffet\footnote{It can be found \hyperlink{https://www.youtube.com/watch?v=ZJzu_xItNkY}{here} or https://www.youtube.com/watch?v=ZJzu\_xItNkY}, he gives his opinion on diversification
\begin{center}
	\textit{''Diversification is protection against ignorance. If you want to make sure that nothing bad happens to you relative to the market, you own everything. There is nothing wrong with that.''}
\end{center}
Assume that he is right and that we have no interest in assets or the financial market what so ever then we should diversify.
If you dont have infinite manpower and aim to own a lot of assets, you wont have time to go investigate the prospects of every asset.

Buffet later continues to argue that its riskier to own fifty stocks in comparison to three.
This 
 
We want to own many assets so that our portfolio is well diversified, forget about it and do other things in life.
However, being ignorant comes at a cost. 
If we want to be ignorant we need some way of constructing portfolios which does not depend on our opinion (or very little of it). 
This thesis presents some ways of doing so which will 

%

%%%%%%%%%%%%%%%
% Give some simple example which we extend and work with.
%%%%%%%%%%%%%%%
\colr{
There are a number of different saying that describe how you should make an important decision. The first that comes to mind is "never make a decision on an empty stomach" or "sleep on it". 
While the former is preemptive and its solution is to keep a healthy diet the former is reactive and stresses you to never make a hasty decision. You should always wait for a while. Deciding what to do with your money or wealth is of course one of these decision. Should you buy a house, a car, save in stocks or simply place everything under the madress and sleep on an uneven surface for the rest of your life? This thesis will not help you with your madress nor will it directly tell you which of these options to pick. However, it does contain some interesting papers on how to allocate your money between a few (or inifinitely many) options!
	
Making sound investments is paramount to portfolio theory. The portfolio is simply a collection of your investments in different assets.
Making optimal allocations is paramount to portfolio theory. 
The term 'optimal' usually refers to the fact that you should not make an investment that does not return any profit and take more risk than necessary. 

Imagine that you are going to invest in a house and borrow the money you need to buy it. 
You have come to an agreement with a bank that you will loan a certain amount and pay interest rate on it. 
The interest is paid on a monthly basis. 
You are very sure that one of the houses that you are looking at will be worth twice the money in two years. 
The decision is simple, if the cost, e.g. the total amount of interest you pay on your loan, is less than the profit you will make selling the house after two years then you should buy the house! 
However, how certain are you that you will be able to sell it at \textit{twice the price in two years}? 
You can not be absolutely certain. 
The housing market might collapse and you will be left with a house that has no worth and no one wants to buy. There is some risk in this decision. 
Lets assume that you are fairly certain that the house will be worth more than twice the money in two years. 
When should we decide to invest in this example? 
It is not as clear as before since we dont know what the house will be worth until we sell it.
The future is always uncertain.

In reality, you dont have a single asset to invest in. 
You most often have many. 
The house is most likely one of many options you can spend your money on. 
You might have a house, kids which you hope will become famous for their skills in sports or stocks. 
Maintaining any of the houses may be costly though it may increase drastically by the investment. 
Driving your kids to practise does not take much investment but it may be highly uncertainty that they will even like it in ten years. 
How should you spend your money on these?
To add to the complexity they might increase (or decrease) in value depending on how you maintain or do not maintain them.
The example is of course valid for any type of \textit{asset} such as stocks, bonds or in essence any type of financial contract. 
Properties might just require more maintainance than a stock.
%Sometimes its worth having everything in a house or another type of \textit{asset} (such as stocks)
The natural question is now, given your collection or \textit{portfolio} of properties, how should you allocate your money when you dont know what the future holds? 
If you think that the market for properties is going to be more profitable than the interest rate you pay for your loans then it makes sense to spread your investment across all properties. 
If the summer house suffers from a mold infection you are most likely fine, your portfolio will not be worthless because you spent money on something that went ..........

 have \textit{diversified} your investment.
Try to do some maintainance on all of your houses in order to spread the \textit{risk} of investing everything in one of them. 
If you do not know what the future holds and . It might even be that some of your other investment is more profitable than you expected and will cover for your loss. In the end you want to minimize the risk or the possibility to loose all your money by diversifying portfolio allocations!

Now comes the issue at hand. 
What happens if you dont know what your assets will be worth in the future?
The first approach could be to specify probabilities for all possible scenarios you think can happen as we did in the example above.
This becomes very unpractical, very quickly. 
The second, more practical approach is to gather information on how your assets have developed in order to get a better estimate of its future values.
This train of thought is the main topic of this thesis. 
We want to construct portfolios which adhere to some criterias while knowing that the future is uncertain and to be honest, we know very little of how our assets will develop in the future. We use data to estimate what we think of the future!

If the dont know what our assets might bring in the future a natural model for it is to assume that they are stochastic. \textbf{Give an example to why portfolios are better than holding a single asset?}
}
%%%%%% ------------------------------------------------------------------------
\chapter{Modern Portfolio Theory}
%%%%%% ------------------------------------------------------------------------
Modern Portfolio Theory was introduced by \cite{markowitz1959portfolio}. 
In his seminal work he argued that any portfolio which simply maximize its profit will result in a naive solution.
Very much like the example in Chapter \ref{ch:intro}. 
The future is unknown and (one of) the best model(s) we have for it is stochastic
Investing all your capital in the asset with the highest return is not sensible if you do not know the future.
Such an investment will cause you to take an extreme amount of risk. 
He therefore argued that any well diversified portfolio should be prefered to any non diversified portfolio. 
Such a portfolio can be obtained through many different procedures but he proposed the use of the first two moments for the allocation problem.
If an asset has high return on average, then that should be accounted for though not at the cost of large amounts of risk. If an asset implies a lot of risk, then we should try to invest small amounts in it.

To state his portfolio allocation problem we first introduce the relevant quantities. In this thesis we never use the asset prices themselves but their a transformation of the relative differences, that is, their simple returns. The simple return is defined as $r_{i,t} := (y_{i,t}-y_{i,t-1})/y_{i,t-1}$ and the log return is then defined as
$
x_{i,t} := \log(r_{i,t} + 1)
$
where $y_{i,t}$ is the price of the $i$th asset at time $t$. We model a portfolio with $p$ assets as $\sum_{i=1}^p w_i x_{i,t} = \bw^\top \bx_t$ where $\bw=(w_1, ..., w_p)$ are the portfolio weights and $\bx_t=(x_{1,t},x_{2,t},..., x_{p,t})$ are the log returns. Notice that this is an approximation. In reality we would want to work with $\sum_{i=1}^p w_i r_{i,t}$ since it is additive in the number of assets. However, logarithmic returns are additive in time which can be desirable. Compounding returns results in simple addition. The difference between the two approaches is very small if the returns are small, which is often true for financial assets. The models we work with in this thesis often rely on the log returns and not the returns. We will omit the time index $t$ to keep the notation tidy unless otherwise stated. 

We assume assets $\bx$ are random with mean $\optn{E}(\bx)=\bmu$ and covariance matrix $\optn{Var}(\bx)=\bSigma$. Although there is usually little restriction on $\bmu$ there is usually very specific restrictions on the covariance matrix $\bSigma$. Since the covariance matrix is a subject of its own we dedicate the next section to it and disregards these restrictions for now. We will merely say that it is well behaved. The restrictions on the mean will be commented on below. Using the two moments for the asset returns the portfolio distribution $x = \bw^\top \bx$ has mean $\optn{E}(x)=\bw^\top \bmu$ and variance $\optn{Var}(x)=\bw^\top \bSigma \bw$. Let $\mu_0$ be the target return that the investor would like to achieve from their portfolio and $\ones$ column vector of ones with appropriate dimensions. \citet{markowitz1959portfolio} considered the following optimization problem
\begin{equation}\label{eqn:markowitz_optim}
\begin{aligned}
& \underset{\bw}{\text{minimize}} 
& & \bw^\top \bSigma \bw \\
& \text{subject to}
& & \bw^\top \ones = 1 \\
& && \bw^\top \bmu \geq \mu_0 \\
&&& w_i \geq 0, i=1,2,..,p
\end{aligned}
\end{equation}
This problem is a quadratic optimization problem with linear equality and inequality constraints. 
The objective is to minimize the variance of the portfolio. 
A natural question is to ask whether or not that implies diversification? As it turns out, minimizing the portfolio variance will encourage diversification.
\textbf{Give example on using convex combinations on variances.} 
 
The constraint $\bw^\top \ones = 1$ essentially states that the investor must invest all available money. 
The weights are scaled according to the amount of cash invested.
The disposition is very different whenever an inequality is used rather than equality. 
As \citet{hult2012risk} states, if $\bw^\top \ones \leq 1$, then the investor could be throwing money away since there is a lot of opportunity left in the market when investing.
The second constraint describes the investors expectations on the portfolio. 
As $\mu_0$ grows, the return of the portfolio will grow. 
However, that has implications for the objective. 
Increasing $\mu_0$ will change the amount of variance the portfolio can achieve. 
\textbf{insert figure for quadratic optimization problem and the different hyperplanes in two dimensions.}

From Figure ... all values of $\mu_0$ are feasible though they may/will come at a cost. 
Depending on the value $\mu_0$ we would be accepting more risk. 
The last constraint is rather simple though have quite large implications. 
It states that the weights can not be negative which means that we can only invest money we have. 
A negative value of $w_i$ in the $i$th asset is called a short position.
You borrow the asset from someone who owns it and then sell it. 
Later on, you can buy it back, hoping that the price is less so that you can make money on such a position. 
For certain types of investors this constraint can be limiting.
We can sometimes obtain less risk if we exclude it, which is the case of all papers in this thesis. That is, this thesis considers
\begin{equation}\label{eqn:mean_variance}
\begin{aligned}
& \underset{\bw}{\text{minimize}} 
& & \bw^\top \bSigma \bw \\
& \text{subject to}
& & \bw^\top \ones = 1 \\
& && \bw^\top \bmu \geq \mu_0 \\
\end{aligned}
\end{equation}
which is what we refer to the mean-variance optimization problem. 
The problem can be solved by langrange multipliers and its solution will almost always result in \textit{active} constraints, e.g. $\bw^\top \bmu = \mu_0$. 
For the interested reader we suggest \citet{} on why that is, we merely take it for granted in this introduction. 
The solution to this problem is very often stated relative to another famous portfolio, namely the Global Minimum Variance (GMV) portfolio and its related quantities (see e.g. \citet{Bodnar2009CaIotEFiEM, bodnar2013equivalence, bauder2018bayesian}). We will continue in the same manner. Let $\bSigma^{-1}$ denote the inverse of $\bSigma$ and 
\begin{equation}
	\bw_{GMV} := \frac{\bSigma^{-1}\ones}{\ones^\top \bSigma^{-1}\ones}, \; R_{GMV} :=\optn{E}(\bw_{GMV}^\top\bx) = \frac{\ones^\top\bSigma^{-1}\bmu}{\ones^\top \bSigma^{-1}\ones}, \;
	V_{GMV} := \optn{Var}(\bw_{GMV}^\top\bx) =\frac{1}{\ones^\top \bSigma^{-1}\ones}.
\end{equation}
 \textbf{explain the quantities.}
  
The solution to the mean-variance problem in \eqref{eqn:mean_variance} is equal to
\begin{equation}\label{eqn:mean_var_solution}
	\bw_{MV} = \frac{\bSigma^{-1}\ones}{\ones^\top \bSigma^{-1}\ones} + \frac{\mu_0 - R_{GMV}}{V_{GMV}} \bQ \bmu,\; \bQ = \bSigma^{-1} - \frac{\bSigma^{-1} \ones \ones^\top \bSigma^{-1}}{\ones^\top \bSigma^{-1} \ones}.
\end{equation}
\textbf{explain the solution. Connect to the two fund theorem?}

The moments of this portfolio is equal to
\begin{equation}\label{eqn:moments_mean_var_solution}
\optn{E}(\bw_{MV}^\top\bx) = R_{GMV} + \frac{\mu_0 - R_{GMV}}{V_{GMV}} \bmu^\top \bQ \bmu, \;
\optn{Var}(\bw_{MV}^\top\bx) =V_{GMV} + \left(\frac{\mu_0 - R_{GMV}}{V_{GMV}}\right)^2 \bmu^\top \bQ \bmu.
\end{equation}
If we set $\mu_0$ equal to $\bmu^\top \bSigma^{-1} \ones / \ones^\top \bSigma^{-1} \ones$ then the portfolio is equal to the GMV portfolio. 
Excluding the constraint $\bw^\top \bmu \geq \mu_0$ all together results in the same solution.
From equation \eqref{eqn:moments_mean_var_solution} we can see that all values $\mu_0$ are rescaled according to the moments of the GMV portfolio. 
If the taget return $\mu_0$ is not $\mu_0>R_{GMV}$ then you are better off with the GMV portfolio in terms of return and risk.
However, if you choose a value $\mu_0>R_{GMV}$ then the portfolio return will be more than $R_{GMV}$.


\textbf{Introduce merton and the efficient frontier and talk about $\bmu^\top \bQ \bmu$.}

%%% ----------------------
\section{Relationship between assets and the (inverse) Covariance matrix}\label{subsec:cov_prec_matrix}
%%% ----------------------
The covariance matrix $\bSigma$ and the precision matrix $\bSigma^{-1}$ are fundemental to mean-variance portfolios. In this section we discuss the restrictions we place on the covariance matrix and what the precision matrix actually represent. 

For a vector $\bx$ with finite second moment, the covariance matrix is defined as $\bSigma=\optn{E}((\bx - \bmu)(\bx - \bmu)^\top)$. 
It contains the variances of each individual element of $\bx$ on the diagonal as well as the covariance between every pair of elements on the off-diagonal. 
That is, each diagonal element corresponds to the univariate case where the variance is equal to $\optn{E}((x_i - \mu_i)^2)$. 
A distribution is usually called degenerate or singular if the variance is equal to zero, which holds true in the multivariate context as well. 
If any diagonal element is zero, then the covariance matrix is called singular.  
However, due to the fact that we involve covariances on the off-diagonal we are forced to work with a broader definition.  
Since we work with real matrices in this thesis, we limit the definition accordingly. From \citet[ch 14.2]{harville1997matrix} we say that a real symmetric $p\times p$ matrix $\bA$ is called 
\begin{itemize}
	\item positive definite if $\bz^\top \bA \bz > 0$
	\item positive semi-definite if $\bz^\top \bA \bz \geq 0$
\end{itemize}
for all nonzero vectors $\bz \in \mathbbm{R}^p$.
In the multivariate case we need to assert that a quadratic form is (strictly) positive in comparison to the univariate setting where we can observe it through the variance. 
Positive- or semi-positive definite can be quite cumbersome to work with. We need to assert that the conditions holds for all vectors $\bz$. 
There are of course many necessary conditions for a matrix to be positive definite though we focus on one of them, using the eigenvalues of a matrix and its eigenvalue decomposition. 
As described in \citet[ch. 21]{harville1997matrix}, an eigenvalue (or characteristic root) $\lambda$ is the solution to 
\begin{definition}\label{def:eigenvalue} 
	Let $\bA$ be a $p\times p$ matrix. The characteristic roots (with multiplicity) are given by the solutions to
	\begin{equation*}
		\left|\bA - \lambda \bI\right| = 0
	\end{equation*}
	where $|\cdot|$ is the determinant of a matrix.
\end{definition} 
Let $\lambda_i$, $i=1,2,...,p$, denote the \textit{ordered} eigenvalues of the matrix $\bA$ such that $\lambda_1\geq \lambda_2 \geq ... \geq \lambda_p$. Given an eigenvalue, the eigenvectors $\bu_i$ are defined by $\bA \bu_i = \lambda_i \bu_i$, $i=1,2,...,p$ where $\bu_i^\top \bu_i = 1$. 
Let $\boldsymbol{\Lambda} = \operatorname{diag}(\lambda_1, \lambda_2,...,\lambda_p)$ and $\bU= (\bu_1^\top, \bu_2^\top, ..., \bu_p^\top)^\top$. 
Using the relation between eigenvalues and their eigenvector we can derive the eigenvalue decomposition of a matrix 

\begin{equation}\label{eqn:eigenvalue_decomp}
	\bA = \bU \boldsymbol{\Lambda} \bU ^\top.
\end{equation}

Now, a necessary condition for a matrix to be positive definite can be directly obtained from the eigenvalue decomposition. 
Let $\bz\in \mathbbm{R}^p$ and $\by := \bU^\top \bz \in \mathbbm{R}^p$, then $\bz^\top \bA \bz = \bz^\top \bU \boldsymbol{\Lambda} \bU ^\top \bz = \by^\top \boldsymbol{\Lambda} \by = \sum_i^p \lambda_i y_i^2$ which is a second degree polynomial. 
If the eigenvalues are all positive then necessarily the matrix is positive definite. 
If there are some eigenvalues which are zero then the matrix is semi-positive definite. 
The eigenvalue decomposition of a matrix $\bA$ contains a lot of information that might not be available at first glance. 
If $\bA$ is a covariance matrix then it contains variances and covariances, describing relations between random variables. 
The eigenvalues are a description of how the system behaves and not individual elements. 
In contrast to the univariate setting, if some eigenvalues are zero then the covariance matrix is singular. 
It depends on variances and covariances.
The eigenvectors are rotations that try to aggregate as much variations as possible along its axis.
Furthermore, if all the eigenvalues are distinct then the eigenvectors are linearly independent (see e.g. \citet[thm 21.4.1]{harville1997matrix}).

In all papers of this thesis we assume that the true covariance matrix is positive definite. 
The assumption has quite a deep economical interpretation.
If one (or more) eigenvalue(s) are zero then there is a possibility to construct a portfolio which does not contain any risk with potentially a positive return. 
An opportunity which should not exist unless the elements of $\bmu$ are all zero.
Assume that $\lambda_p=0$, let $\bu_p$ be its eigenvector and $\bw = \bu_p / \sum_i^p u_{ip} $. The variance of that portfolio is zero since all eigenvectors are orthonormal and its mean is $\bw^\top \bmu$ which can be non-zero unless the elements of $\bmu$ are all zero. 
If the true covariance matrix is not positive definite there might exist arbitrage opportunities, e.g. the possibility of making profit without taking any risk.

The inverse covariance (or precision) matrix $\bSigma^{-1}$ is a very important object for MPT. 
From the mean-variance problem \eqref{eqn:markowitz_optim} we can derive its solution, which was stated in \eqref{eqn:mean_var_solution}. 
While the covariance makes is presence in \eqref{eqn:markowitz_optim} the solution contains the precision. 
Let $\bB$ be a symmetric $p \times p$ matrix. It is an inverse of the matrix $\bA$ if $\bB \bA = \bI$. 
We claim that $\bB = \bU \boldsymbol{\Lambda}^{-1} \bU^\top$ is a valid inverse which is of course true since,
$$
\bB \bA = \bU \boldsymbol{\Lambda}^{-1} \bU^\top \bU \boldsymbol{\Lambda} \bU^\top
$$
\begin{itemize}
	\item Talk a little about the inverse and how that can look, give an example?
\end{itemize}
% Connect to the papers!

% Connect to the papers!

%%%%%% ------------------------------------------------------------------------
\chapter{Statistical models and estimation}
%%%%%% ------------------------------------------------------------------------
To \textit{pratically} use the portfolios described by \eqref{eqn:mean_var_solution} and all its variations.

\begin{itemize}
	\item Multivariate Normal
	\item Wishart and inverse wishart distribution, properties
	\item Matrix variate location and scale. 
\end{itemize}

\begin{definition}
	A random vector $\bx \in \mathbbm{R}^p$ follows a multivariate normal distribution with mean vector $\bmu \in \mathbbm{R}^p$ and positive definite covariance matrix  $\bSigma \in \mathbbm{R}^{p \times\, p}$ if its density is given by 
	\begin{equation}\label{eqn:multi_density}
	\frac{|\bSigma|^{-1/2}}{2\pi} \exp \left\{-\frac{1}{2} \left(\bx - \bmu \right)^\top\bSigma^{-1}\left(\bx - \bmu \right) \right\}
	\end{equation}
	where $|\bB|$ is the determinant of the matrix $\bB$. We usually use the notation $\bx \sim N_p(\bmu, \bSigma)$ to indicate the above.
\end{definition} 

\begin{definition}
	The random matrix $\bS$ of size $p \times p$ follows a $p\times p$ dimensional Wishart distribution with $n$ degrees of freedom, $n > p$, if its density is given by
	\begin{equation}\label{eqn:wishart_density}
	\frac{|\bS|^{(n-p-1)/2} |\bSigma|^{- n/2} }{2^{pn/2} \Gamma_p (n/2) } \exp\left\{-\frac{1}{2} \operatorname{tr}(\bSigma^{-1}\bS)  \right\}
	\end{equation}
	where $ \Gamma_p (\cdot) $ is the multivariate gamma function and $\operatorname{tr}(\cdot)$ is the trace operator, i.e. the sum of the diagonal elements and $\bSigma, \bS$ are both positive definitie. We use the notation $\bS \sim W_p(n, \bSigma)$ to indicate that $\bS$ follows a Wishart distribution with the given parameters.
\end{definition}

\begin{definition}
	A positive definitie random matrix $\bA$ is said to be distributed according to a $p\times p$ dimensional inverse Wishart distribution with $n$ degrees of freedom and positive definitie parameter matrix $\bV$ if its density is given by
	\begin{equation}\label{eqn:inverse_wishart}
	\frac{2^{-(n-p-1)p/2} |\bSigma|^{(n-p-1)/2} }{\Gamma_p ((n-p-1)/2) |\bA|^{n/2}} \exp\left\{ -\frac{1}{2} \bA \bV \right\}, \; n> 2p
	\end{equation}
	which we denote $\bA \sim W^{-1}_p(n, \bV)$ to indicate that $\bV$ follows an Inverse Wishart distribution with the given parameters.
\end{definition}


\begin{lemma}\label{lem:stoc_rep}
	$\bx$ is a $k$ dimensional elliptically contoured distributed random variable with mean $\bmu$ and dispersion matrix $\bSigma$ where $\text{rk} (\bSigma) = p$, if and only if 
	$$
	\bx \stackrel{d}{=} \bmu + r \bV \bu
	$$
	where $r >0$ is a random variable, $\bu$ is a random vector which is uniformly distributed on the sphere, in $\mathbbm{R}^k$, $\bV\bV^\top=\bSigma$, and $r$ and $\bu$ are independent.
\end{lemma}

\section{Estimation - the name of the game}

\begin{itemize}
	\item What are the implications of using $\bS$ instead of $\bSigma$?
	\item why is $\bS$  always an admissible estimator? (MM)
	\item Other types of estimators and why they might be better than $\bS$.
\end{itemize}

\begin{remark}
	Unconditional and conditional covariance estimation. Prediction is very hard and constructing viable models. Are returns predictable? Should we even try?
\end{remark}

\section{Simulations, inverses and why stochastic representations are valuable}
\begin{itemize}
	\item Motivating simulations and the issue with inversions.
	\item Simulation of multi- or matrixvariate distributions can be very computationally consuming.
	\item ...
\end{itemize}
%%%%%% ------------------------------------------------------------------------
\chapter{Higher dimensions or portfolios with many assets}
%%%%%% ------------------------------------------------------------------------
In the previous chapter we presented different ways of estimating the covariance matrix. Under certain conditions and/or statistical models, the sample covariance matrix inherited certain properties. 
%If we hold $p$ constant and let $n$ grow the sample covariance matrix is consistent. 
If we have a lot of data on the assets that we are trying to invest in then we can be certain that we will hold the correct portfolio. It estimates the correct object of interest. However, if we believe in diversification then $p$ should be big as well. It should decrease the risk (variance) of the portfolio. We then venture in the curse of dimensionality. By introducing one new asset to a portfolio of size $p$ we need to estimate the variance and all covariances which is equal to $p+1$ quantities. In terms of estimation uncertainty, this does not scale well. To understand the effect of estimation uncertainty in high dimensional portfolios we make use of the vast subject of Random Matrix Theory (RMT, see e.g.  \citet{bai2010spectral} INTRODUCTION MATERIAL) 

\section{Random Matrix Theory and eigenvalues}
The subject of RMT has many applications. It was originally developed in the context of statistical physics by \citet{} WIGNER. The theory has developed and is now used in many fields such as combinatorics , computational biology, wireless communication and finance. This section gives a brief introduction and will then cover a small part of the field more indepth, namely high dimensional limits of traces involving the inverse sample covariance matrix and transformations thereof. 

To understand 

\begin{itemize}
	\item Theorem 1 of \citet{rubio2011spectral}.
\end{itemize}


Definition \ref{def:pos_def_matrix} can be quite complicated to work with. It states that we need to check all different vectors $\bz$ to assert if a matrix is positive definite. A potentially easier way to deduce if a matrix is positive (semi-)definite is to use the eigenvalue decomposition. The eigenvalue decomposition comes from the following eigenvalues 

\section{The implications of large dimensions for portfolio theory}
\begin{itemize}
	\item Give small example of the different quantities of the 
\end{itemize}
%%%%%% ------------------------------------------------------------------------
\chapter{Summary of Papers}
%%%%%% ------------------------------------------------------------------------
The papers presented here are among a total of ... papers produced. These are selected based their common theme.
\section*{Paper 1}
Sampling dist.
\section*{Paper 2}
TP portfolio CSN model.

From Chapter \ref{ch:intro} we constructed an example with the possibility of lending money. That is usually referred to as a risk-free rate or asset, commonly denoted denoted $r_f$. It can be a bond or the interest rate on a loan. However, from \eqref{eqn:markowitz_optim} there is no risk-free rate. To introduce this option into our portfolio we simply add the risk-free rate as $w_0 r_f + \bw^\top \bx$ and optimize over $w_0$ as well. However, since $w_0 r_f + \bw^\top \ones=1$ we can introduce the following problem
\textbf{Introduce a figure of the efficient frontier with the riskfree rate}
\section*{Paper 3}
This paper deals with the fact that taking limits changes estimates. If we invest in the GMV portfolio today and wait a week the estimates in the portfolio will have changed. Assuming that we are ok with rebalancing the portfolio at fixed time points, say every month or year depending on the investment horizon, we develop a reweighting scheme to minimize movement. 

The paper is accompanied by a R package, available on CRAN. The package implements the functionality and provides it through a simple interface for any developer/investor to use. Give short intro to the package. 

\section*{Paper 4}
Double shrinkage?
%\section{Paper 5}
% Backtesting in higher dimensions, many VaR predictions and many
%%%%%% ------------------------------------------------------------------------
\chapter{Future research}
%%%%%% ------------------------------------------------------------------------
There are many possible extensions and future projects to the theory at hand. 

\begin{itemize}
	\item Are shrinkage intensities for the sample covariance matrix optimal for the precision or MPT problem? 
	\item One of the most interesting issues of the elliptical distribution and its inverse sample dispersion matrix. What are the moments of $(\mathbf{Z} \mathbf{R} \mathbf{Z}^\top)^{-1}$? 
	\item There are different ways of incorporating estimation uncertainty, one solution is robust optimization. Are there connections to be made? Is robust optimization just Emperical Bayes?
	\item Sequential reweighting extension to Paper 3. When should we reweight?
	\item Stochastic representations and markov chains. Simple to sample from stochastic representations, are there ways to boost performance of MCMC methods?
	\item BEKK models are usually hard to fit and use for MPT. Their forecasts are not always positive definite. The first issue is simple to cope with if one can formulate the models as Recurrent Neural Networks and use the efficient deep-learning libraries Torch or Tensorflow. By doing so, one would also get the possibility to develop new models by constructing new layers to the networks. It would also be easier to integrate different sources of information in the models.
\end{itemize}

Although not part of this Phd thesis Flipped classroom and online learning tools.

\bibliography{references}
\end{document}
