---
title: "Optimal portfolios in the high-dimensional setting"
author: "Erik Thorsén"
email: erik.thorsen@math.su.se
header-includes: |
  \newcommand{\bSigma}{\mathbf{\Sigma}}
  \newcommand{\bS}{\mathbf{S}}
  \newcommand{\bw}{\mathbf{w}}
  \newcommand{\ones}{\mathbf{1}}
  \newcommand{\bb}{\mathbf{b}}
  \newcommand{\bI}{\mathbf{I}}
output:
  ioslides_presentation:
    widescreen: true
    incremental: true
    logo: SU_logo_CMYK.png
    self_contained: true
    css: ["https://cdn.jsdelivr.net/npm/bootstrap@5.1.3/dist/css/bootstrap.min.css", "styles.css"]
bibliography: ../references.bib
---

```{r setup, include=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(kableExtra))
knitr::opts_chunk$set(echo = FALSE)
options("kableExtra.html.bsTable" = T)
```

# Introduction

## Development of wealth 

```{r, fig.cap= "", fig.align='center'}
set.seed(124)
n<-100
tibble(
  "Low risk - Low reward" = c(1, cumprod(1+rnorm(n,0.003, 0.02))),
  "High risk - High reward" = c(1,cumprod(1+rnorm(n, 0.01, 0.06))),
  "x"=0:n
) %>% 
  pivot_longer(-x) %>% 
  ggplot() +
  geom_line(aes(x=x, y=value, color=name), size=1.05) +
  theme_void() +
  scale_color_brewer("", palette="Dark2")
```

## The ''Low risk - Low reward'' portfolio

Low risk portfolio from @markowitz1959portfolio is the Global Minimum Variance (GMV) portfolio
$$
\min_\bw \bw^\top \bSigma \bw,\\\;\; \text{s.t.} \;\bw^\top \ones = 1.
$$
where the solution is
$$
\bw_{GMV}=\frac{\bSigma^{-1} \ones}{\ones^\top \bSigma^{-1} \ones}.
$$

## This thesis in a slide - Estimation & diversification

- In practice we do not know $\bSigma$, _estimate_ with $\bS$ from $n$ obs. How accurate is that estimate?
- We want $p$ to be large, e.g _diversify_ a lot.
  * $\bS$ has $p(p+1)/2$ parameters $\rightarrow$ curse of dimensionality!
  * We want to use $\bS^{-1}$ for GMV, not $\bS$. How accurate is $\bS^{-1}$ or
  $$
  \widehat{\bw}_{GMV} = \frac{\bS^{-1} \ones}{\ones^\top \bS^{-1} \ones}?
  $$
- Estimates makes everything complicated $\rightarrow$ understand difference between using $\bSigma$ and $\bS$ or construct ''better'' estimates?

## Summary of papers in this thesis 

```{r}
tibble(
  "Paper" = as.roman(1:5),
  "Aim" = c("Understanding uncertainty", "Understanding uncertainty", "Better estimates", "Better estimates", "Better estimates"),
  "Portfolios"=c("Many!", "TP", rep("GMV", 3 )),
  "Model" = c("Normal", "Normal with skewness", rep("Location-scale",3)),
  "$n,p < \\infty$" = c("$\\cdot$", "$\\cdot$", "", "", ""),
  "$n,p\\rightarrow \\infty$" = "$\\cdot$",
) %>% 
  kbl(escape = FALSE, align = c("c", "c", "c", "c", "c")) %>% 
  kable_styling(bootstrap_options = "striped",  full_width = F, position = "center", font_size = 25)
```
<br/>

- Paper I & II tries to understand what happens when we use estimates instead of true parameters.
- Paper III-V constructs ''better estimates'' for the portfolio weights.

# Paper I & II | Together with Taras Bodnar, Nestor Parolya, Holger Dette,<br/> Stepan Mazur & Farrukh Javed  {#title}

## Paper I & II - Sampling distributions & understanding uncertainty

- If we use $\bS$ instead of $\bSigma$ we introduce estimation uncertainty.
- Aim is to answer questions such as: 
  * Have we bought the right amount of units in asset A? 
  * We think that our portfolio should grow $3\%$ each year, how certain are we about it? 
- To do this we assume a lot, but can get a deeper understanding of what happens! 

## Paper I & II - Results {.build}

*Paper I*

- Returns follow a normal distribution and observations are iid.
- Investigates the sampling distribution of _any_ portfolio in the MPT framework.
  * Sample quantities are biased, not even consistent
  * We construct consistent estimators! 

*Paper II*

- Returns follow a skew-normal distribution, still iid.
- Investigates the sampling portfolio from the expected quadratic utility/''mean-variance'' problem _with a risk-free asset_
  * Same thing as the above, bias present due to skewness 

# Paper III, IV & V | Together with Taras Bodnar & Nestor Parolya 

## Shrinkage in high-dimensional portfolios

- Even though $\widehat{\bw}_{GMV}$ is supposed to minimize risk, it wont if $p$ is close to $n$.

- These papers consider portfolios on the following form
$$
\widehat{\bw} =\psi \widehat{\bw}_{GMV} + (1-\psi)\bb
$$
where $\bb$ is "some portfolio".
  * Introduces bias to the portfolio but will lower its variance.
  * How do we determine $\psi$? $$\min_{\psi} \widehat{\bw}^\top \bSigma \widehat{\bw}$$
_note that $\bSigma$ is present again!_

## Paper IV - Is variance informative?

```{r, warning=FALSE, fig.height=2.5, fig.align='center', fig.pos='top'}
offset <- 0.01
tibble(x = 1:11,
       y = 1) %>% 
  ggplot() +
  geom_line(aes(x=x, y=y), linetype=13, 
            arrow = arrow(length=unit(0.20,"cm"), ends="last", type = "closed")) +
  # period 1
  geom_segment(aes(x=1, xend=5, y=1-offset, yend=1-offset, color="Period 1"), size=3) +
  geom_text(x=3, y=1-offset*3/2, label=expression(n[1])) +
  # period 2
  geom_segment(aes(x=5, xend=8, y=1+offset, yend=1+offset, color="Period 2"), size=3) +
  geom_text(x=6.5, y=1+offset*3/2, label=expression(n[2])) +
  coord_cartesian(ylim=c(1-offset*5, 1+5*offset)) +
  scale_color_brewer("", type="qual", palette = 6) +
  theme_void()
```

- Comparison between three different strategies to estimate $\psi$: @frahm2010, @bodnar2018estimation & sample GMV
- Compared by the empirical out-of-sample variance
$\hat{V}_{n_1, n_2} = \widehat{\bw}_{n_1}^\top \bS_{n_2} \widehat{\bw}_{n_1}$
and the empirical relative loss
$$
\hat{L}_{n_1, n_2} = \left(1-\frac{p}{n_2}\right)\hat{V}_{n_1, n_2} \ones^\top \bS^{-1}_{n_1} \ones -1
$$

## Paper IV - Results {.build}

- The Variance is informative, though bad for some $\bSigma$.
- Dominance results/Ordering in loss and variance of the three different portfolios $n,p \rightarrow \infty$.
- Empirically ordering holds $n,p < \infty$, example $n_1=n_2$

```{r, fig.height=3.5, fig.width=10,cache=TRUE, fig.align='center'}
s <- 19
date <- Sys.Date()
### Load data
load("SP500_returns.dat")
colnames(returns) <- str_remove(colnames(returns), ".Close")
df_wide <- returns %>%
  as.data.frame() %>%
  rownames_to_column("Date") %>%
  mutate(Date = lubridate::as_date(Date))

seed <- 123; p_ <- 190; n_window <- 200;
# omit Date
ports <- sample(2:(ncol(df_wide)), p_)

df_wide_adj <- zoo::na.locf(df_wide, fromLast = TRUE)
dates <- df_wide_adj$Date

# Extract p x n matrix
Y_all <- df_wide_adj %>%
  select_at(ports) %>%
  as.matrix()
# set up the target portfolio
b <- rep(1/p_, p_)
df_moving_window_overlap <- map_dfr((n_window+1):(length(dates)-n_window), ~{
  # extract data from period  
  Y <- Y_all[(.x-n_window):(.x), ] 
  Y_eval <- Y_all[(.x+1):(.x+n_window), ] 
  n <- nrow(Y)
  #  
  S <- var(Y)
  S_inv <- solve(S)
  # weights
  o <- rep(1, nrow(S))
  w_gmv <- S_inv %*% o / as.numeric(t(o) %*% S_inv %*% o)
  # FM
  alpha_FM <- 1- ((p_-3)/(n-p_+2)) / (
    as.numeric(t(o) %*% S_inv %*% o) * as.numeric(t(b) %*% S %*% b) -1 
  )
  w_fm <- alpha_FM*w_gmv + (1-alpha_FM)*b
  # BPS
  alpha_BPS <- (
    (1-p_/n)*(
      (1-p_/n)*as.numeric(t(o) %*% S_inv %*% o) * as.numeric(t(b) %*% S %*% b)-1)
  )/(
    p_/n + (1-p_/n)*(
      (1-p_/n)*as.numeric(t(o) %*% S_inv %*% o) * as.numeric(t(b) %*% S %*% b)-1)
  )
  w_bps <- alpha_BPS * w_gmv + (1-alpha_BPS)*b
  #######
  m <- nrow(Y_eval)
  S_out <- var(Y_eval) 
  S_out_inv <- solve(S_out)
  c(
    "L_GMV (Sample)" = ((1-p_/m))*as.numeric(t(o) %*% S_out_inv %*% o) * 
      as.numeric(t(w_gmv) %*% S_out %*% w_gmv) - 1,
    "L_FM" = ((1-p_/m))*as.numeric(t(o) %*% S_out_inv %*% o) * 
      as.numeric(t(w_fm) %*% S_out %*% w_fm) - 1,
    "L_BPS" = ((1-p_/m))*as.numeric(t(o) %*% S_out_inv %*% o) * 
      as.numeric(t(w_bps) %*% S_out %*% w_bps) - 1,
    "V_Traditional"=as.numeric(t(w_gmv) %*% S_out %*% w_gmv),
    "V_FM"=as.numeric(t(w_fm) %*% S_out %*% w_fm),
    "V_BPS"=as.numeric(t(w_bps) %*% S_out %*% w_bps), 
    "p"=p_
  )
})
df_moving_window_overlap  %>%
  group_by(p) %>%
  mutate(date=dates[(2*n_window+1):length(dates)]) %>%
  pivot_longer(cols = `L_GMV (Sample)`:V_BPS) %>%
  separate(name, sep = "_", into=c("type", "name")) %>%
  mutate(type=if_else(type == "L", "Loss", "Variance")) %>% 
  filter(type == "Loss") %>% 
  ggplot() +
  geom_line(aes(x=date, y=value, color=as.factor(name)), size=0.73) +
  theme_minimal() +
  scale_color_brewer("Weights type", type="qual", palette = 6) +
  theme(legend.position = "right", text = element_text(size=12))+
  labs(title="Relative loss (lower is better)", subtitle=expression('190 random stocks from S&P500, ' ~ n[1] ~'='~ n[2]~ '= 200 with a moving window.'))
```


## Paper III - Dynamic Shrinkage estimation

```{r, warning=FALSE, fig.height=2.5, fig.align='center', fig.pos='top'}
offset <- 0.01
tibble(x = 1:11,
       y = 1) %>% 
  ggplot() +
  geom_line(aes(x=x, y=y), linetype=13, 
            arrow = arrow(length=unit(0.20,"cm"), ends="last", type = "closed")) +
  # period 1
  geom_segment(aes(x=1, xend=5, y=1-offset, yend=1-offset, color="Period 1"), size=3) +
  geom_text(x=3, y=1-offset*3/2, label=expression(n[1])) +
  # period 2
  geom_segment(aes(x=5, xend=8, y=1+offset, yend=1+offset, color="Period 2"), size=3) +
  geom_text(x=6.5, y=1+offset*3/2, label=expression(n[2])) +
  # period 3
  geom_segment(aes(x=8, xend=10, y=1-offset, yend=1-offset, color="Period 3"), size=3) +
  geom_text(x=9, y=1-offset*3/2, label=expression(n[3])) +
  coord_cartesian(ylim=c(1-offset*5, 1+5*offset)) +
  scale_color_brewer("", type="qual", palette = 6) +
  theme_void()
```


- We work with $\widehat{\bw}_{n_i}=\psi_{n_i} \widehat{\bw}_{GMV;n_i} + (1-\psi_{n_i})\widehat{\bw}_{n_{i-1}}$ (nontrivial extension of @bodnar2018estimation)
- Estimate $\psi_{n_i}$ by $\hat{\psi}_{n_i}$ at each transition, equal/optimal when $n,p \rightarrow \infty$
- Datasets can be overlapping, so Period 2 can contain data from Period 1!


## Paper III - Results

- We compare a number of different strategies, most notably @lw20 nonlinear covariance estimator
- Simulations show that the method is robust and often provide the smallest relative loss.
- Empirically, @lw20 is better at taking ''low-risk'' positions, our method provides better out-of-sample variance, mean, turnover.
- Methods are in the `DOSPortfolio` package (on CRAN!).

## Paper V - ''Double'' shrinkage

- An extension to the linear shrinkage, includes ridge type regularization to $\bS$
$$
\widehat{\bw} = \psi \frac{(\bS + \eta \bI)^{-1}\ones}{\ones^\top(\bS + \eta \bI)^{-1}\ones} + (1-\psi) \bb
$$

- $c \in [0,\infty)$ or $p>n$, previously $c \in [0,1)$ or $p<n$.
- We cant determine $\eta, \psi$ analytically $\rightarrow$ estimator of $\widehat{\bw}^\top \bSigma \widehat{\bw}$, consistent when $n,p \rightarrow \infty$.
- Have to assume that $\bSigma$ have bounded spectral norm

## Paper V - Results {.build}

- We compare to a number of different strategies, @lw20, @bodnar2018estimation, @frahm2010 for different $\bb$
- Simulations show that ''Double'' is also robust and on par with the @lw20.
- Empirically, Double performs well, example $p=431, n=250$
```{r}
tibble(
  "Variance" = c(0.0121, 0.01217, 0.01088, 0.01117, 0.01116, 0.01105 , 0.01322), 
  "Mean"=c(0.000437, 0.00042, 0.000496, 0.000516, 0.000512, 0.000515, 0.000375), 
  "Turnover" = c(3165.55, 3192.65, 69.99, 5.19, NaN, 660.56, 4254.45),
  "Strategy" = c("BPS_ec", "BPS_ew", "Double_ec","Double_ew", "ew", "LW2020", "GMV (sample)")
) %>% 
  mutate(across(c(Turnover, Variance), ~{
          case_when(
            is.nan(.x) ~ "",
            .x == min(.x, na.rm=T) ~ paste0("$\\mathbf{", .x, "}$"),
            .x == nth(.x, n=2, order_by = .x) ~ paste0("$\\mathit{", .x, "}$*"),
            TRUE ~ paste0("$", .x, "$")
          )}),
         Mean = case_when(
           is.nan(Mean) ~ "",
            Mean == max(Mean, na.rm=T) ~ paste0("$\\mathbf{", Mean, "}$"),
            Mean == nth(Mean, n=2, order_by = desc(Mean)) ~ paste0("$\\mathit{", Mean, "}$*"),
            TRUE ~ paste0("$", Mean, "$")
         )) %>% 
  pivot_longer(Variance:Turnover) %>% 
  pivot_wider(names_from = Strategy,
              values_from = value) %>% 
  set_names(c("", "ec", "ew", "ec", "ew", "ew", "LW2020", "GMV (sample)")) %>% 
  kable(caption = "Variance and Turnover, lower is better. Mean, higher is better.",
      booktabs = T, escape = FALSE, longtable = T) %>% 
  kable_styling(font_size = 14, bootstrap_options = c("striped")) %>%
  add_header_above(c(" "=1, "BPS" = 2, "Double" = 2, " "=3))
```

- _Double is consistently the best in terms of low-risk positions!_

## Summary {.build}

This thesis provides some ways to better understand the investment process or to provide ''good'' portfolios when $n,p\rightarrow \infty$. 
We either

- assume a lot to get a deeper understanding of the quantities of interest.
- assume little, construct ''good'' portfolios but we might provide less intuition.

All code for this thesis is available on [my github page (https://github.com/Ethorsn)](https://github.com/Ethorsn/Phd-thesis)

## References
