Early in my graduate studies I asked a professor what it takes to complete a PhD and he answered grit.
The purpose of a PhD is to spend a number of years intensively digging deeper into a subject, coming up with ideas that most often have been solved, does not seem to be solvable or cannot be solved at all.
It is a very humbling experience and there are many people that I would like to thank for making it a fun one as well!

The first person to thank is my supervisor Taras Bodnar.
I am immensely grateful for your guidance, knowledge and discussions over the years.
Thank you for giving me the opportunity of doing this.
I could not have asked for a better supervisor.
I also want to thank my co-supervisor Joanna Tyrcha for your support.

I want to give special thanks to my co-authors that also made this an exciting and great learning experience.
Nestor Parolya, whom I have gotten the pleasure to work extensively with over the last few years. 
You are a co-author of almost all the papers in this thesis and it has been a joy.
I have learned a lot from our discussions.
Thank you, Stepan Mazur and Farrukh Javed for the fun discussions and great collaborations!
Furthermore, thank you Vilhelm Niklasson, Mathias Lindholm and Holger Dette for the (very) long and rewarding discussions! 

Thank you Ola Hössjer for letting me be a part in developing and teaching the course Statistical Data Processing.
I have learned so much from the experience and grown as a person from it!

I also want to thank all my colleagues at the Department of Mathematics.
It has been a incredibly fun place to work and I think that the environment has been, and will continue to be, great.
There are a number of individuals that have made my experience even better.
Thank you Vilhelm, Gustav and Stanislas for being wonderful people to share an office with.
I hope to continue our brewing adventures Vilhelm and the wonderful banter Gustav!
I also want to thank Felix, Måns, Sebastian and Hampus whom I have gotten to spend a lot of time with during the COVID-19 pandemic. 
It made everything so much easier.

Last but not least, I cannot express enough love and gratitude towards my family. 
Mom, dad and my beloved sister, I cannot thank you enough for all your support.
Johanna, you make me a much happier person and I look forward to many more years with you and our son Gabriel. 