The following papers are included in this thesis.
% \section*{Papers included in this thesis}
% \printbibliography[keyword={included_in_thesis}, title={Papers included in this thesis}, heading=subbibliography]
\begin{enumerate}[P{A}PER I: ]
  \setlength{\itemsep}{3.3mm} %Set the vertical distance between the items
  \item \fullcite{bodnar2020sampling_thesis} \label{earlier_version}
  \item \fullcite{javed2021tangency_thesis}
  \item \fullcite{bodnar2021dynamic_thesis}
  \item \fullcite{bodnar2021empirical_thesis}
  \item \fullcite{bodnar2022double_thesis}
\end{enumerate}
\textbf{Author contribution:} 
E. Thors\'en has actively contributed to developing the content of all papers.
In the first paper, E. Thors\'en contributed with the derivation of the theoretical results and with writing the majority of the paper in collaboration with all other authors. The original idea was presented by T. Bodnar.
In the second paper, E. Thors\'en contributed with the theoretical derivations and writing the majority of the paper in collaboration of S. Mazur and F. Javed. The original idea was presented by S. Mazur and F. Javed.
The original idea of the third paper was born from discussions with T. Bodnar and N. Parolya. 
In this paper, E. Thors\'en contributed with the theoretical derivations and the writing of the paper.
The original idea of the fourth paper stemmed from discussions from the third paper together with T. Bodnar and N. Parolya.
In this paper, E. Thors\'en contributed with the theoretical derivations and the writing of the paper.
The fifth paper also sprung from the third and fourth paper. 
In this paper, E. Thors\'en contributed to the theoretical derivations, the writing of the paper and the idea behind it.
E. Thorsén carried out all computer implementations in all articles.
In all other aspects the authors contributed equally.

\textbf{General comment:} An earlier version of Paper \ref{earlier_version} was included in the Licenciate thesis of E. Thors\'en, \citet{thorsen2019assessment}.

\nocite{bodnar2021quantile_thesis, BODNAR2022127120, bodnar2021bayesian_thesis, DOSPortfolio_thesis}
\printbibliography[keyword={papers_list}, heading=subsubbibliography]
